---
title: "Data Release Report: 2009 Yellow-billed Loon Survey in Cape Krusenstern National Monument and Bering Land Bridge National Preserve, Alaska"
output:
  pdf_document:
    toc: false          # Include table of contents
    # number_sections: true  # Number sections
    # fig_caption: true     # Enable figure captions
    # latex_engine: xelatex # Use xelatex engine (optional)
  html_document:
    toc: true
    toc_float: true
---

```{r setup,echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Setup
options(knitr.kable.NA = '')
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(odbc)
library(leaflet)

# Build a database connection
Connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inp2300irmadb01\\ntwk", Database = "AKRO")
DeliverablesSchedule = dbGetQuery(Connection,"SELECT Identifier, Deliverable, Description FROM vwDeliverablesSchedule2 WHERE (ProtocolID = 35) ORDER BY Identifier")
```

```{r echo=FALSE}

dir = r'(C:\Work\VitalSigns\ARCN Western Yellow-billed Loons\Data\2009)'
setwd(dir)


# Use the below for sequentially numbering tables and figures
QCCheckCounter = 1
TableCounter = 1
FigureCounter = 1
```


Scott D. Miller\
Information Technology Specialist/Data Manager\
National Park Service, Arctic Inventory and Monitoring Network\
240 W. 5th Ave\
Anchorage, AK 99501

`r format(Sys.Date(), "%B %d, %Y")`

# Abstract

```{r,echo=FALSE}
odata = read.csv(file.path(dir,'YB06 2009 YBLO Survey Occupancy Dataset.csv'))
# glimpse(odata)
pdata = read.csv(file.path(dir,'YB06 2009 YBLO Survey Productivity Dataset.csv'))
# glimpse(pdata)

OStartDate = min(odata$Date,na.rm=TRUE)
OEndDate = max(odata$Date,na.rm=TRUE)
PStartDate = pdata %>% filter(Date != '') %>% summarize(min(Date,na.rm=TRUE))
PEndDate = max(pdata$Date,na.rm=TRUE)

Abstract = paste0("The National Park Service conducted an aerial survey of Yellow-billed Loons in Cape Krusenstern National Monument and Bering Land Bridge National Preserve, Alaska. An occupancy survey was flown from ",OStartDate," to ",OEndDate," and a productivity survey was flown from ",PStartDate," to ",PEndDate,". This Data Release Report describes the deliverables published from this survey.")

```

`r Abstract`

# Dataset Sensitivity Statement

**[SENSITIVE DATASET]** NPS internal use only.

The resource described in this report has been designated as a species
of concern and is legally protected. The rules for dissemination for
Yellow-billed loons specifically state (Koelsch, 2019):

*"Legally Protected Under National Parks Omnibus Management Act, Section
207, of 16 U.S.C. § 5937; Specific location (geographic coordinates) of
this species of concern will not be released to the general public due
to their global rarity and nest site fidelity. Any such data collected
will be redacted from the publicly released datasets. This protects
collaborating agencies data as well; but does not preclude sharing
between collaborating agencies. Processed data with coordinates redacted
(such as summaries, processed maps, or products) remain publicly
available. This will be done indefinitely."*

# Table of Contents

\tableofcontents

# Access

The data release associated with this report is available from the NPS
Data Store at the following URL:
<https://irma.nps.gov/DataStore/Reference/Edit/2315500>.

# License

U.S. Government Works Data and content created by government employees
within the scope of their employment are not subject to domestic
copyright protection under 17 U.S.C. § 105. Government works not
specifically designated for withholding by law are, by default, in the
U.S. Public Domain.

# Deliverables

Data deliverables For this monitoring program are defined In Flamme
M.J., J.H. Schmidt, J. Mizel and S.D. Miller. 2018. Yellow-billed loon
monitoring protocol for the Arctic Network: Protocol narrative version
1.0. Natural Resource Report. NPS/ARCN/NRR—2018/1738. National Park
Service. Fort Collins, Colorado, Version 1.00, IRMA reference code
2256435. This protocol Is available through the National Park Service's
Integrated Resource Management Applications Data Store.

## Deliverables Schedule

```{r,echo=FALSE}
knitr::kable(DeliverablesSchedule)

```

## Deliverable Specifications

### YB01 Survey geodatabase (YBLO.gdb)

Geodatabase containing processed QA/QC-checked survey data, including
feature classes/layers of the loon and nest data. There is a different
geodatabase for each year.

### YB02 Park Observer files

Folder containing iPad Park Observer .pov files and affiliated
geodatabase containing the raw, field collected data prior to importing
and processing those data in the GIS.

### YB03 Waypoints

Waypoints collected by the pilot or observer during a loon survey

### YB04 Data collection forms, booklets and field notebooks

Annotated data collection forms in data booklets and field notebooks
generated during loon survey.

### YB05 Certified database snapshot

A snapshot of the loons monitoring certified dataset for publication.

### YB06 Observations data file

Deliverable YB06 consists of the digitized loons data collected during a
survey. These data should be in comma separated values text file format.

### YB07 Tracklogs

Deliverable YB07 consists of the digitized tracklog data collected
during a survey. These data should be in comma separated values text
file format.

### YB08 Lakes dataset

Deliverable YB08 consists of a spatial data file containing the surveyed
lakes polygons. These data should be in comma separated values text file
format.

### YB09 Plots dataset

Deliverable YB09 consists of a spatial data file containing the surveyed
Plot polygons. These data should be in comma separated values text file
format.

### YB10 Data collection forms

Deliverable YB10 consists of physical paper data collection forms, if
applicable.

### YB11 Photographs

Deliverable YB11 consists of photographs taken during a survey.

# Summary of Datasets Created

```{r, echo=FALSE}
# Set the directory path
directory_path = r'(C:\Work\VitalSigns\ARCN Western Yellow-billed Loons\Data\2009)'

# List all files in the directory (non-recursive)
files <- list.files(path = directory_path, full.names = TRUE)

# Obtain file info including last modification time
file_info <- file.info(files)

# Build a data frame with filenames and timestamps
result_df <- data.frame(
  Filename = basename(files)                  # Extract just the filenames, without path
  #Description = ''                 # Last modification time
)

# Print the resulting data frame
knitr::kable(result_df %>% filter(Filename != 'Misc') %>% arrange(Filename))

```


# Data Quality Evaluation

The data within the data records listed above have been reviewed by staff in the NPS Inventory and Monitoring Division to ensure accuracy, completeness, and consistency with documented data quality standards, as well as for usability and reproducibility. This dataset is suitable for its intended use as of the date of processing (`r format(Sys.Date(), "%B %d, %Y")`).

# Methods

See Flamme, et al. (2018) for methods.


# References

Flamme, M.J., J.H. Schmidt, J. Mizel and Miller, S.D. 2018. Yellow-billed loon monitoring protocol for the Arctic Network: Protocol narrative version 1.0. Natural Resource Report. NPS/ARCN/NRR—2018/1738. National Park Service. Fort Collins, Colorado [https://irma.nps.gov/DataStore/Reference/Profile/2256435](https://irma.nps.gov/DataStore/Reference/Profile/2256435)

Koelsch J. 2019. To: Wald EJ. Re: Handling of Protected Natural Resource
Data - BELA <https://irma.nps.gov/DataStore/Reference/Profile/2263477>
