---
title: ""
format:
  typst:
    toc: false
    columns: 1
---

```{r setup, include=FALSE, echo=FALSE,results='hide'}

# Set up report parameters here
SurveyYear = 2007  # Set this to the desired report year. 

dir = paste0(r'(C:\Work\VitalSigns\ARCN Western Yellow-billed Loons\Data\)',SurveyYear)

knitr::opts_chunk$set(warning = FALSE, message = FALSE) # Prevent overwhelming messages from reaching user

options(download.file.method = "wininet")
# Make kable show blanks for NAs
options(knitr.kable.NA = '')

#install.packages("sqldf")
# Load libraries
library(odbc)
library(ggplot2)
library(ggtext)
library(tidyverse)
library(leaflet)
library(leaflet.extras)
library(sf)
library(janitor)
library(english)

# ODBC connection to the  monitoring SQL Server Database. 
# Contact data manager for server and database details.
Connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna", Database = "ARCN_Loons")

# Current date
CurrentDate = format(Sys.Date(), format='%B %d, %Y') # Get the current date

# Function to convert a date string to a long date
GetLongDate <- function(Date) {
  return(format(as.Date(Date), format="%B %d, %Y"))
}

# These counters are used to number the tables and figures. Each gets incremented after use.
TableCounter = 1
FigureCounter = 1

```

```{r,echo=FALSE,name='Get the data',results='hide'}
# Get the survey details
Sql = paste0("SELECT [Year]
      ,[SurveyName]
      ,[DeliverablesReferenceCode]
      ,Convert(Varchar(4000),[Abstract]) as Abstract
      ,Convert(Varchar(4000),[Notes]) as Notes
      ,[RecordInsertedDate]
      ,[RecordInsertedBy]
      ,[ReportReferenceCode]
      ,[Personnel]
      ,[ProtocolVersion]
      ,[DataQualityNotes]
  FROM [ARCN_Loons].[dbo].[Surveys] WHERE [Year] = ",SurveyYear)
SurveyDetails = dbGetQuery(Connection,Sql)


# Get the entire survey dataset
SurveyDataSql = paste("SELECT  [Year]
      ,[Park]
      ,[Plot]
      ,[Lake]
      ,[Date]
      ,[Species]
      ,[ScientificName]
      ,[CommonName]
      ,[Loons]
      ,[Nests]
      ,[Pairs]
      ,[Chicks]
      ,[Eggs]
      ,[Location]
      ,[Activity]
      ,[IsIncidentalObservation]
      ,[IsNest]
      ,[IsPair]
      ,[SurveyType]
      ,[LakeType]
      ,[Latitude]
      ,[Longitude]
      ,[Notes]
      ,[ObservationID]
  FROM [ARCN_Loons].[dbo].[Dataset_SurveyData]
  WHERE [Year] = ",SurveyYear,"
  ORDER BY [Year],Plot,Lake,[Date]",sep="")

# Get the survey data
data = dbGetQuery(Connection,SurveyDataSql)

# Publish a dataset, if needed
if(1==0){
  YB06 = dbGetQuery(Connection,paste0("SELECT * FROM Dataset_Certified WHERE Year = ",SurveyYear))
  YB06File <- paste0(dir,"/YB06 ",SurveyYear," ARCN Loons Survey Data.csv")
  write.csv(YB06,YB06File,quote=TRUE,na="",row.names = FALSE)
}


# Spatial layers, Parks and Alaska
# Parks = st_read('https://irma.nps.gov/DataStore/DownloadFile/704836')
#AK = st_read('https://irma.nps.gov/DataStore/DownloadFile/746052')
GISPath = r'(C:\Work\GIS Common Layers\GCS\)'
Parks = st_read(paste0(GISPath,'AKParks_reprojected.shp'))
AK = st_read(paste0(GISPath,'AK_GCS.shp'))

# Get the Plots spatial polygons from the database
Plots <- st_read(
  Connection,
  query = "
    SELECT 
      Year, Plot, Park, PlotType, 
      PlotFeature.STAsBinary() AS geom
    FROM Plots
    WHERE Year = 2025
    ORDER BY Plot",
  geometry_column = "geom"
)


# Group and summarize the loons survey data
ResultsSummary <- data %>%
  filter(IsIncidentalObservation != 1 | is.na(IsIncidentalObservation)) %>%
  group_by(`Common name` = CommonName) %>%
  summarize(
    `Loons` = sum(Loons, na.rm = TRUE),
    Pairs = sum(IsPair, na.rm = TRUE),
    Nests = sum(IsNest, na.rm = TRUE),
    Eggs = sum(Eggs, na.rm = TRUE),
    Chicks = sum(Chicks, na.rm = TRUE),
    #`Eggs/nest` = ifelse(is.nan(sum(Eggs, na.rm = TRUE)/sum(IsNest, na.rm = TRUE)), 0, sum(Eggs, na.rm = TRUE)/sum(IsNest, na.rm = TRUE)),
    #`Chicks/nest`= ifelse(is.nan(sum(Chicks, na.rm = TRUE)/sum(IsNest, na.rm = TRUE)), 0, sum(Chicks, na.rm = TRUE)/sum(IsNest, na.rm = TRUE)),
    n = n()
  )

# Set the CRS (e.g., WGS84 or Alaska Albers)
Plots <- st_set_crs(Plots, 4326)  # or 3338 if using Alaska Albers

```

```{r echo=FALSE}
# Generate an abstract
# The abstract is stored in the database Surveys table, but it does not always exist. If it exists, use it in the report, otherwise generate one on the fly (consider updating the database with it afterwards)
GeneratedAbstract = paste0(
"National Park Service (NPS) biologists conducted an aerial survey for loons over "
,as.character(english(as.numeric(data %>% summarize(n_distinct(as.Date(Date))))))," flying days from "
,GetLongDate(as.character(data %>% summarize(as.character(min(as.Date(Date),na.rm=TRUE)))))," to "
,GetLongDate(as.character(data %>% summarize(as.character(max(as.Date(Date),na.rm=TRUE)))))," in the Seward Peninsula and Cape Krusenstern area of 
Western Alaska. Biologists surveyed ",
english(data %>% filter(Species=='YBLO') %>% summarize(Plots = n_distinct(Plot,na.rm=TRUE)) %>% pull(Plots))," 12km x 12km Yellow-billed Loon monitoring 
plots and observed "
,ResultsSummary %>% summarize(Loons=sum(Loons)) %>% pull(Loons)," loons of "
,english(data %>% filter(Species != 'UNLO') %>% summarize(nSp = n_distinct(Species,na.rm=TRUE)) %>% pull(nSp))," species, "
,data %>% summarize(nPairs = sum(IsPair,na.rm=TRUE))," nesting pairs, "
,data %>% summarize(nNests = sum(IsNest,na.rm=TRUE))," nests, "
,data %>% summarize(nEggs = sum(Eggs,na.rm=TRUE))," eggs in "
,data %>% summarize(n = n())," observations. ",
"In terms of Yellow-billed loons, the primary loon of interest to the NPS, biologists observed "
,data %>% filter(Species == 'YBLO') %>% summarize(nLoon = sum(Loons,na.rm=TRUE))," loons, "
,data %>% filter(Species == 'YBLO') %>% summarize(nPairs = sum(IsPair,na.rm=TRUE))," nesting pairs, "
,data %>% filter(Species == 'YBLO') %>% summarize(nNests = sum(IsNest,na.rm=TRUE))," nests, "
,data %>% filter(Species == 'YBLO') %>% summarize(nEggs = sum(Eggs,na.rm=TRUE))," eggs in "
,data %>% filter(Species == 'YBLO') %>% summarize(n = n())," observations."
)


# Execute this Update query against the database to update the Surveys.Abstract attribute for this survey record.
#UpdateAbstractSQL = paste0("UPDATE Surveys Set Abstract = '",gsub("'","''",GeneratedAbstract),"' WHERE Year = ",SurveyYear)

# Determine which abstract to use, the database one, or a generated on the fly one.
if(is.na(SurveyDetails$Abstract) == TRUE | SurveyDetails$Abstract == ''){
  Abstract = GeneratedAbstract
}else{
  Abstract = SurveyDetails$Abstract
}


```

# Yellow-billed Loon Survey Summary, `r SurveyYear`

## National Park Service

### Arctic Inventory and Monitoring Network

Scott D. Miller\
U.S. Department of the Interior\
240 W. 5th Ave Anchorage\
Alaska 99501

`r CurrentDate`

# Abstract

`r Abstract`

**Keywords:** Yellow-billed Loon, Loon, Monitoring, Population, Abundance, Alaska, Productivity, Occupancy, *Gavia adamsii*, coastal, `r SurveyYear`

# Introduction

The National Park Service monitors the abundance and productivity of loons in Bering Land Bridge National Preserve (BELA) and Cape Krusenstern National Monument (CAKR) in western Alaska. 
This report summarizes the results of a Yellow-billed Loon *(Gavia adamsii)* aerial survey conducted in `r SurveyYear` by the Arctic Inventory and Monitoring Network (ARCN) in and around 
Bering Land Bridge National Preserve and Cape Krusenstern National Monument.

Yellow-billed Loons are a species of concern due to their small global population. Their breeding range spans the arctic tundra regions of Russia, Canada, and Alaska. 
In Alaska, nesting is primarily restricted to the Arctic Coastal Plain (north of the Brooks Range) and western Alaska, particularly on or near the Seward Peninsula. 
This survey contributes to the NPS' long-term loons monitoring dataset, providing valuable insights into the abundance of Yellow-billed Loons in western Alaska.

# Methods

NPS biologists followed ARCN's established protocol and standard operating procedures ([Flamme et al., 2018](https://irma.nps.gov/DataStore/Reference/Profile/2216938)) in this survey
and readers are referred to that document for detailed descriptions of methods used. NPS surveyed all lakes \>7 ha in size in the study area using the U.S. Fish and Wildlife Service's 
lake-circling methods and established random plots for this aerial survey to optimize detection of Yellow-billed Loons in their nesting habitats. 
Two pilot-observers teams in tandem fixed-wing aircraft conduct repeat surveys of the plots scattered across the two park units. 
Flights were conducted at an altitude of less than 500 feet and all five species of loons encountered (Yellow-billed, Common, Red-throated, Pacific and Arctic) and their nests are recorded.

## Survey Dates

```{r,echo=FALSE,tab.cap=paste("Table ",TableCounter,". Survey dates.",sep="")}
# Survey dates
SurveyDatesSummaryDF = data %>% summarize(
    Start = min(Date,na.rm=TRUE),
    End = max(Date,na.rm=TRUE),
    Days = n_distinct(Date,na.rm=TRUE)
  )
SurveyDatesSummary = paste0("The ",SurveyYear," loon survey took place over ",SurveyDatesSummaryDF$Days," days from ",GetLongDate(SurveyDatesSummaryDF$Start)," to ",GetLongDate(SurveyDatesSummaryDF$End),".")
# knitr::kable(t(SurveyDatesSummary))
# TableCounter = TableCounter + 1
```
`r SurveyDatesSummary`

## Survey Area

The survey area includes 13,572 km2 of potential Yellow-billed Loon habitat in portions of BELA and CAKR in western Alaska.

```{r, echo=FALSE,fig.cap=paste0("Figure ",FigureCounter,". Map of the ",SurveyYear," aerial loon survey area showing Cape Krusenstern National Monument (coastal park in green, north) and Bering Land Bridge National Preserve, in green, south) with standard 12x12km monitoring plots as gray squares.")}

# Get bounding box of the Plots layer
bbox <- st_bbox(Plots)
buffer = 0.1

# Plot stuff
ggplot() +
  
  # Alaska polygon
  geom_sf(data = AK, fill = "white", color = "gray80", alpha = 0.5) +
  
  # Parks polygons
  geom_sf(data = Parks, fill = "palegreen4", color = NA, alpha = 0.2) +

  # Plot polygons
  geom_sf(data = Plots, fill = NA, color = "gray40", alpha = 0.3,size=0.1) +
  geom_sf_text(data = Plots, aes(label = Plot), size = 3, color = "gray40") +

    # Park labels
  annotate("text", x = -164.2, y = 67.6, label = "Cape Krusenstern\nNational Monument",size = 3, color = "black", angle = 0,fontface='bold') +
  annotate("text", x = -164.5, y = 65.8, label = "Bering Land Bridge National Preserve",size = 3, color = "black", angle = 0,fontface='bold') +
  
  # Bounding box restriction
  coord_sf(
    xlim = c(bbox["xmin"] - buffer, bbox["xmax"] + buffer),
    ylim = c(bbox["ymin"] - buffer, bbox["ymax"] + buffer),
    expand = FALSE
  ) +
  
  # Theming
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "lightsteelblue1"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(title = paste0(SurveyYear," Loon Survey Study Area"),
    x='Longitude',
    y='Latitude')

FigureCounter = FigureCounter + 1
```

# Results

## Observations Summary

A summary of loon observations during the `r SurveyYear` survey appears in Table `r TableCounter`.

```{r,echo=FALSE,tab.cap=paste("Table ",TableCounter,". Summary of loon observations.",sep="")}


# Print results table
ResultsSummary %>%
  adorn_totals() %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  knitr::kable()

TableCounter = TableCounter + 1
```

## Distribution By Species

The distribution of loon observations and nest locations  by species during the `r SurveyYear` loon survey in BELA and CAKR are presented below.

```{r,echo=FALSE,include=FALSE}

GetLoonDistributionMap2 = function(LoonSpecies,FigureNumber){
  # Plot loon observations
  MapData = data=data %>% filter(Species == LoonSpecies & Species != 'UNLO')
  
  # Get the common and sci names of the LoonSpecies parameter
  CommonName = MapData[1,'CommonName']
  ScientificName = MapData[1,'ScientificName']
  
  # Make a map title
  MapTitle = paste0(CommonName," Observations")
  
  # Make a map caption
  FigureCaption = paste0("Figure ",FigureNumber,". Distribution of ",CommonName," (*",ScientificName,"*) observations.\nLoon sightings shown as small black triangles and nest locations as blue squares.")

  # Build a map of loon observations
  map = ggplot(MapData) +
    
    # Alaska and Park polygon layers
    geom_sf(data = AK, fill = "white", color = "gray80", alpha = 0.5) +
    geom_sf(data = Parks, fill = "palegreen4", color = NA, alpha = 0.2) +
  
    # Plot polygons
    geom_sf(data = Plots, fill = NA, color = "gray80", alpha = 0.3,size=0.1) +
    geom_sf_text(data = Plots, aes(label = Plot), size = 3, color = "gray70") +
    
    # Observations
    geom_point(aes(x=Longitude,y=Latitude),color='black',size=0.5,shape=17) +
    
    # Nests
    geom_point(data = data %>% filter(Species == LoonSpecies & IsNest == 1), aes(x=Longitude,y=Latitude),color='royalblue1',size=1,shape=15) +
    
    # Label the parks
    annotate("text", x = -164.2, y = 67.6, label = "Cape Krusenstern\nNational Monument",size = 3, color = "gray40", angle = 0) +
    annotate("text", x = -164, y = 65.8, label = "Bering Land Bridge National Preserve",size = 3, color = "gray40", angle = 0) +
    
    # Bounding box restriction
    coord_sf(
      xlim = c(bbox["xmin"] - buffer, bbox["xmax"] + buffer),
      ylim = c(bbox["ymin"] - buffer, bbox["ymax"] + buffer),
      expand = FALSE
    ) +
    
    # Labels
    labs(
      title = MapTitle,
      caption = FigureCaption) +
    
    # Theming
    theme_minimal() +
    theme(
      panel.background = element_rect(fill = "lightsteelblue1"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.caption.position = "plot", 
      plot.caption = element_markdown(hjust = 0)
    )
  
  return(map)
}
# GetLoonDistributionMap2('YBLO',33)
# GetLoonDistributionMap2('RTLO',34)

```

```{r,echo=FALSE,results='asis'}
SpeciesList <- data %>%
  filter(Species != 'UNLO') %>%
  distinct(Species) %>%
  arrange(Species) %>%
  pull(Species)  # Extracts as a character vector

for (Sp in SpeciesList) {
  # Print the loon species name as a header
  CommonName = as.character(data %>% filter(Species == Sp) %>% distinct(Species,CommonName) %>% select(CommonName))
  cat("  \n",sep="")
  cat("### ",CommonName,"  \n\n",sep="")
  cat("The distribution of ",CommonName," observations and nest locations are shown in Figure ",FigureCounter,", below. Observations are shown as black triangles and nest locations as blue squares.  \n",sep="")
  cat("  \n",sep="")
  cat("  \n",sep="")
  # Print the loon distribution map
  print(GetLoonDistributionMap2(Sp,FigureCounter))
  cat("  \n",sep="")
  
  # Increment the figure counter
  FigureCounter = FigureCounter + 1
}

```

# References

Flamme, M.J., J.H. Schmidt, J. Mizel and S.D. Miller. 2018. Yellow-billed loon monitoring protocol for the Arctic Network: Protocol narrative version 1.0. Natural Resource Report. NPS/ARCN/NRR—2018/1738. National Park Service. Fort Collins, Colorado <https://irma.nps.gov/DataStore/Reference/Profile/2256435>



```{r,include= FALSE}
# Copy the report to the loon directory
src =  paste0(getwd(),"/ARCN Loons Survey Summary Report.pdf")
dst <- paste0(dir,"/",SurveyYear," ARCN Loons SurveySummary Report.pdf")
dst2 <- paste0(getwd(),"/",SurveyYear," ARCN Loons SurveySummary Report.pdf")
cat("Source file : ",src," exists: ",file.exists(src))
if (file.exists(src)) {
  file.copy(src, dst, overwrite = TRUE)
  file.copy(src, dst2, overwrite = TRUE)
}

```
